Sequence Specific Modeling of Cell-Free Protein Synthesis

### 2017-03-21 Paper_v1.tex
new figs and text updates

\subsection{Model derivation and validation}
The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
A schematic of the metabolic network, which consisted of 270 reactions and 145 species, is shown in Fig.~\ref{fig:network}A.
Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
different cell-free \textit{E.~coli} extracts. The cell-free metabolic network, model parameters, model code,
and each protein sequence are available in the supplemental materials.

Cell-free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values.
On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
The cell-free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.

Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_network.pdf}
\caption{Sequence specific flux balance analysis. A. Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway. Thick gray arrows indicate withdrawal of precursors for amino acid synthesis. B. CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption. Error bars denote the standard deviation of experimental measurements. C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption. Error bars denote a 10\% coefficient of variation. D. Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
\label{fig:network}
\end{figure}
These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.

@@

\subsubsection{Productivity}
The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prod}).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prod}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The difference in the T7 and  P70a promoters did not alter the qualitative productivity performance of CAT.
The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
A further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Productivity.pdf}
\caption{CFPS productivity of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Productivity across the ensemble (error bars represent 95\% CI). B. Mean productivity versus carbon number. Single trendline (dotted line) calculated across all cases (R\textsuperscript{2} = 0.99). Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Prod}
\end{figure}
%The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation, as well as synthesize each amino acid required for the production of the protein of interest.

\subsubsection{Energy efficiency}
The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 30\% when amino acids were not available (Fig.~\ref{fig:Energy}).
Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
When amino acids were removed from the media, protein production resulted in a low energy efficiency; below 30\% depending on the protein. 
This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis, in addition to being available for energy generation.
We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
The same inverse trend was observed as for productivity, except that it was linear.
The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, leading to a lower energy efficiency of protein synthesis.
The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information).
Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 25\% regardless of carbon number.
In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.3 $\pm$ 1.4\% compared to the theoretical maximum of 72 $\pm$ 7.1\%.
This shows CFPS systems have a lot of room for improvement: first, the experimental setup still produced certain amino acids; these reactions could be turned off.
Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.

%This may be a potential area for improvement in CFPS to optimize for energy utilization.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
\caption{CFPS energy efficiency of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency (Error bars represent the 95\% CI of the ensemble). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.97) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.79). Dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Energy}
\end{figure}

@@

The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system.

@@

For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 56.9 $\pm$ 2.4\% for CAT.
Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 57.0 $\pm$ 2.2\% for CAT.

@@

Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
In the experimentally constrained case, CAT was produced with a carbon yield of 6\% compared to the theoretical maximum of 57\%.

@@

In addition, ssFBA is solved by setting protein production as the objective function to maximize.

@@

\includegraphics[width=1.00\textwidth]{./Figures/CarbonYield.pdf}
\caption{CFPS carbon yield of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency across the ensemble (error bars represent the 95\% CI). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.92) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.83). Dagger: constrained by experimental measurements and excluded from trendline.}

@@

Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
A single trendline was formulated for the cases where amino acids were supplied in the media (first two cases) since they had similar performance, while another trendline was formulated for when amino acids were not available (available in Supporting Information).
As the protein size increased, the carbon yield decreased, suggesting that large proteins may be less feasible for cell-free production.

@@

\subsection{Sensitivity analysis}
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation by ribosomes being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis in that case.
In addition, oxygen uptake had more significance in the third case than in the first two cases, since in that case oxygen was required to synthesize amino acids and power the system.
Also, oxygen determined how effectively glucose was utilized; therefore, if glucose uptake rate effects productivity, then oxygen could be expected to have a similar effect.
On the other hand, amino acid uptake showed significance for the cases where amino acids were available, and was even higher for the second case (without AA synthesis), as it was the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}

When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all three cases while the sensitivity to transcription/translation parameters decreased slightly.
The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate was an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
Oxygen uptake showed significant importance for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
Across all three cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency and carbon yield of deGFP.
In the first two cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
Glucose uptake was only seen to be significant when amino acids were not available in the media (third case), since it was the only source of carbon for protein synthesis and energy generation.
%This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.
We also investigated the sensitivity of carbon yield to network fluxes; it followed the same trends as the energy efficiency sensitivity analysis.
It is unknown how active oxidative phosphorylation is compared to in \textit{in~vivo} systems.
To investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
Interestingly, oxidative phosphorylation has a strong effect on the carbon yield in all cases.
The first two cases where amino acids are supplied followed the same trend: ranging from a carbon yield of 20\% to 55\%, depending on the oxidative phosphorylation activity.
Once amino acids were removed from the media the carbon yield dropped to about 10\%, but still reached the same maximum of 55\% with higher oxidative phosphorylation flux.
When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid synthesis.
In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
It is unclear how active oxidative phosphorylation is in CFPS, since the reactions rely on electron transport from membrane vesicles.
The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
%Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
\includegraphics[width=0.7\textwidth]{./Figures/Yield_Ox.pdf}
\caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
\label{fig:oxphos_yield}

>>>>

\subsection{Model derivation and validation}
The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
different cell-free \textit{E.~coli} extracts. The cell-free metabolic network, model parameters, model code,
and each protein sequence are available in the supplemental materials.

Cell-free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values.
On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
The cell-free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_network.pdf}
\caption{Sequence specific flux balance analysis. A. Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway. Thick gray arrows indicate withdrawal of precursors for amino acid synthesis. B. CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption. Error bars denote the standard deviation of experimental measurements. C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption. Error bars denote a 10\% coefficient of variation. D. Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
\label{fig:network}
\end{figure}

Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.

@@

\subsubsection{Productivity}
The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prod}).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prod}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h. 
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Productivity.pdf}
\caption{CFPS productivity of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Productivity across the ensemble (error bars represent 95\% CI). B. Mean productivity versus carbon number. Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99). Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Prod}
\end{figure}

\subsubsection{Energy efficiency}
The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Energy}).
Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein. 
This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
The same inverse trend was observed as for productivity, except that it was linear.
%The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 28\% regardless of carbon number.
In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
This shows CFPS systems have a lot of room for improvement: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potential could be diverted towards a protein of interest.
Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.

%This may be a potential area for improvement in CFPS to optimize for energy utilization.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
\caption{CFPS energy efficiency of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency (Error bars represent the 95\% CI of the ensemble). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -1.65$\cdot$10\textsuperscript{-2}x+90.95; R\textsuperscript{2} = 0.91) and trendline for AA synthesis without uptake (grey dotted line; y = -3.18$\cdot$10\textsuperscript{-3}x+33.48; R\textsuperscript{2} = 0.78). Dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Energy}
\end{figure}

@@

%The cases supplemented with amino acids in the media showed the highest carbon yield.
There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.

@@

For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.

@@

%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.

@@

In addition, ssFBA is solved by maximizing the flux through the protein production reaction.

@@

\includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
\caption{CFPS carbon yield of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency across the ensemble (error bars represent the 95\% CI). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for AA synthesis without uptake (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90). Dagger: constrained by experimental measurements and excluded from trendline.}

@@

%Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
%The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}B).
The trendlines showed good predictability for estimating carbon yield for a certain range of potein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.

@@

\subsection{Sensitivity analysis}
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for CAT, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield which followed the same trend of significance as energy efficiency (available in Supporting Information).
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}

When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all cases while the sensitivity of the transcription/translation parameters decreased.
The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
Oxygen uptake was the most significant for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
Across all cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency of CAT synthesis.
In cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
Glucose uptake was only important when amino acids were not available, since it was the only source of carbon for protein synthesis and energy generation.
%This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.
It is unknown how active oxidative phosphorylation is compared to in \textit{in~vivo} systems.
To investigate this further we compared CAT carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
Interestingly, oxidative phosphorylation had a strong effect on the carbon yield in all cases.
The cases where amino acids were supplied followed the same trend: ranging from a carbon yield of 20\% to 58\%, depending on the oxidative phosphorylation activity.
Once amino acids were removed from the media, the carbon yield dropped to about 10\%, and reached a maximum of 52\%.
When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid biosynthesis.
In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
Glucose can not be fully oxidized and therefore fermentation pathways are used. 
Oxidative phosphorylation relies on electron transport from membrane vesicles, however CFPS has no cell membrane, thus it is expected CFPS has limited oxidative phosphorylation activity.
The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
%Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
\begin{figure}[t!]
\includegraphics[width=0.7\textwidth]{./Figures/Ox_yield.pdf}
\caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
\label{fig:oxphos_yield}
\end{figure}
